Preregistration for modelling in ecolgy and conservation

### 2020-09-16 index.Rmd
Complete first draft abstract. Rework iterative model development section. rework adaptive preregistration and model checking sections.

---
title: "Preregistration for modelling in ecolgy and conservation"
description: |
  Working Draft.
author:
  - name: Elliot Gould 
    affiliation: School of BioSciences, University of Melbourne
  - name: Chris Jones
    affiliation: Department of Environment Land and Water
  - name: Libby Rumpff
    affiliation: School of BioSciences, University of Melbourne
  - name: Hannah Fraser
    affiliation: School of BioSciences, University of Melbourne
  - name: Fiona Fidler
    affiliation: School of BioSciences, University of Melbourne
date: "`r Sys.Date()`"
bibliography: ./data/references.bib
output: 
  distill::distill_article:
    self_contained: false
---
**Intended journal:** Methods in Ecology and Evolution.

@@

*The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:*

*Point 1: set the context for and purpose of the work;*

*Point 2: indicate the approach and methods;*

*Point 3: outline the main results;*

*Point 4: identify the conclusions and the wider implications.*

**Keywords** *(no more than 8 for MEE).***:** *preregistration, preregistration template, modelling, applied ecology, conservation decision-making, metaresearch, reproducibility, non-hypothesis testing(?).*

@@

Ecology as a discipline is at risk of a "reproducibility crisis" (Fidler et al. 2017), with rates of Questionable Research Practices (QRPs) in ecology similar to those in other fields (Fraser et al. 2018), such as psychology and medicine, where the crisis has received considerable attention, and provoked heated discussion among researchers within the broader scientific community. In particular, there are serious problems with the way ecological models and the modelling process are reported in the published literature. However, concerns about poor model reporting practices are not new, and have been raised repeatedly --- nearly a decade ago researchers concluded that most model reporting was neither transparent or comprehensive [@Schmolke:2010fd; @deVos:2011], and that a culture of model reporting was yet to be established [@Grimm:2014es]. Moreover, most models were assessed predominantly on the basis of scientific originality during peer-review  [@Schmolke:2010fd] to the effect that the burden of evaluating models for credibility is placed onto the reader or model-user, rather than the peer-review process, especially in the case of extremely complex models [@Alexandrov:2011iv].

Despite attempted to promote thorough and transparent model documentation using 'Good Modelling Practice' (e.g. [@Jakeman:2006ii;@Alexandrov:2011iv]), more recent quantitative assessments of transparency and reproducibility demonstrate that the transparent model reporting is still an ongoing problem. Just as Jakeman argued that many models in the published literature lack a clear statement of model objectives and purpose, lack descriptions of model assumptions and implications, and also fail to report model results, including validation and evaluation in 2006, [-@Jakeman:2006ii], [@Feng2019]'s quantitative assessment of Ecological Niche Models found that reporting of model evaluation and model results was less than adequate.

@@

The modelling workflow + adaptive preregistration process described below articulates the inner logic of the preregistration templates that underpins the content, structure of the template, as well as our proposed procedure and interface for implementing the template.

@@

Table X, below, describes the 'model scientific workflow' underpinning the template, which was identified through the structured collaborative workshop, literature review and qualitative coding. By articulating this workflow, we explicitly state the workflow and sets of activities that the template presumes. We identified and coded decision-points into 6 'phases' in the process of developing, evaluating and analysing a model or models in ecological application: 'problem formulation', 'define conceptual model', 'formalise and specify model', 'model validation and evaluation', 'model analysis', 'model application'. These phases are comprised of a series of smaller 'steps' and 'substeps', which are defined and described in detail within Table X. The broad phases and steps are used to structure the template by grouping similar modelling activities together, and the descriptions and definitions of the phases and steps are used as 'help-text' in the template interface.

We synthesised the literature on many different environmental modelling

We synthesised modelling workflows from various

Although there are many different environmental modelling frameworks

We synthesised the literature on many different environmental modeling frameworks with the outputs of our workshop.

\-

@@

@@

During our structured discussions, we identified the iterative nature of model development as a key barrier to adopting preregistration.

\<Insert refs to literature describing this\>.  \<Explain what I mean by iterative model development and provide examples\>

\<Explain why this is a barrier to preregistration\>

Typically, the process of preregistration is completely distinct from and precedes the implementation of the analysis plan laid out in the preregistration. In fact, preregistration has been defined as "the action of confirming an unalterable version of one's research plan prior to collecting data" (Nosek et al. 2018) or prior to analysing the data (Ref, or link to secondary data preregistration template). This results in a linear research workflow, whereby a researcher shifts from the process of ideation and design, documentation of that design in a preregistration, to the process of action, where the researcher executes the plan specified in the preregistration \<insert link to this fig..\>

Translating the preregistration process to a modelling research context is thus challenging because the iterative nature of model development conflicts with the inner logic of existing preregistration templates that presume a linear research workflow. For instance, some modelling decisions are inherently dependent on the outcomes and results of previous decisions and analyses, which involves conducting some preliminary or investigatory analyses before being able to specify future decision-steps in an analysis plan, contravening the critical(?adj) feature of preregistration of specifying the analysis plan prior to seeing the data. Moreover, some items in the template, particularly at later stages of modelling, might not be able to be answered by the researcher until the model is fully or close to fully specified on a first-pass (e.g. specifying exactly what sensitivity analyses or uncertainty analyses will be conducted). This might especially hold true for more complex process models. \<Mention that this is not a flaw in preregistration, but rather reflects the inner logic of preregistration for hypothesis-testing contexts\>

__The *checking* problem__

Authors, reviewers, editors and readers may all engage in the process of checking a report or manuscript against its preregistration to verify that the study and analyses were conducted as specified in the preregistration. However, existing preregistration formats \<do not encourage/make the process of checking difficult\> --- static text-based documents are designed for authors to quickly input the desired information. In fact, *Pu et al* go so far as to say that this format "verge[s] on being write-only media" and are therefore completely ill-suited to the checking task.

The task of model checking is further complicated by the non-linear and iterative nature of model development. It is difficult to verify the conducted study and analysis against the preregistered analysis for existing preregistration formats used in hypothetico-deductive research . However, when the modelling and preregistration is iterative, such that there might be interim preregistrations, then this means that there are *multiple* versions of the preregistration to check against the completed analysis.

The question we faced then is, how do we design both the format of the template, and its procedure for completion by the author, so that we facilitate preregistration checking --- especially when this task is complicated by the iterative nature of model development? *Pu et al* put forth two design elements that may facilitate checking. They recommend designs that "make relevant preregistration easier to *query*" and that "encourage wider *coverage* of preregistration content during review".

For *Pu et al.* The process of *querying* a preregistration consists of "paper-to-preregistration" and "preregistration-to-paper" lookup whereby the reader searches for relevant information from the paper in the preregistration and *vice-versa*. This requires that the report of paper must be specific in referring to parts of the preregistration, *and*, that the author has tagged the relevant part of the preregistration in their report or manuscript.

Consequently we propose a methodology for managing the preregistration process that leverages features of GitHub that implement the task of linking the actual study and analysis that was conducted to the pre-specfied plan in the preregistration \<link to part of GH methodology that explains how checking is accounted for in the procedural design\>. In addition, we designed our shiny interface for completing the preregistration to include a 'checking mode' that easily allows the author to state how and why the the analysis might have diverged from the planned \<link to part of shiny app interface that discusses checking mode\>.

---------

*I'm not exactly sure whether this content should go here in a summarised form or whether it should go into its respective places coming up in the manuscript and we just link to it.*

*How does our GH procedure incorporates checking:*

1\. Using issues to track committed work on the same analysis task, and that issue is spawned from a template item so the conducted work is inherently linked to that initial plan.

2\. In the preregistration document

*How does our shiny-preregistration Interface incorporates checking:*

\- Checking mode - The author can switch to checking mode after having conducted the analysis for a given PRT item. *Or* after the preregistration is complete.

What do you think about this process? What if we allowed check boxes in our preregistration software design so that the author could switch to 'checking mode' and they can tick some boxes Yes / No / Partial match. For No and Partial, they briefly describe \_how\_ it varies from the preregistration, and why (could link to the GitHub issue thread where conversation and decisions were logged during the process of conducting the analyses).

This is going to take some further thought because there might be \_interim\_ versions of the document. Our process means that the author keeps adapting the preregistration until they actually go ahead and implement it... so doesn't this mean that they will \_always\_ follow their latest version of the preregistration?? I guess then checking entails a comparison of first and interim preregistrations against the actual conducted analysis, not just the final.

AND it is possible that in other cases the author deviated from their plan for some reason, and didn't follow our proposed process using GH. So we want to allow for this to occur, but for the author to be able to state \_that\_ this happened, \_how\_ it differs, and why.

---------

@@

@@

The process of preregistration as it is currently implemented is tailored towards the hypothetic-deductive model of the scientific method, and thus follows a linear process, where the process of planning and preregistering is distinct and completely precedes the execution of that plan. The researcher shifts from ideation, to study/analysis design and preregistration, to collecting and analysing the data, writing the report or manuscript, to publishing ([*cos.io/rr*](http://cos.io/rr)) \<See also Figure 1 in Pu et al.\> The iterative nature of model development, however, inherently precludes this model of preregistration from being applied to model-based research in ecology and conservation.

Firstly, as Srivastava notes, there are some decisions that are either difficult to anticipate, or simply cannot be made in advance \<Srivastava\>. This is particularly true for decisions occurring at the later phases of the modelling construction and development process (Fig X), where downstream decisions depend on the outcomes of earlier decisions and outcomes of modelling analyses. For example, in the modelling process we have described (previous section), some decisions cannot be made unless you have a fully specified model, such as \<sensitivity analysis, and other model evaluation tasks and test\>.

Secondly, very often in modelling, preliminary or exploratory analyses must be implemented before decisions can be made about how to proceed. For example, modellers might need to check the distribution of particular variables in order to determine how to specify the model, perform assumption checks, or check for collinearity or spatial autocorrelation. Sometimes the results of these analyses and checks mean that the modeller is forced to return to earlier decision-points and change the analysis. For instance, the model might fail to converge, or perhaps there is insfufficient data to fit the desired model such that the model runs out of variation to apportion to all the different variables \<there's a name for this\>.

Consequently, we take an 'expanded view of preregistration', that fits with the concept of 'adaptive preregistration' proposed by \<Srivastava\>. This view of preregistration breaks with the current model of preregistration, wherein the author writes a single deterministic preregistration containing a rule for every decision, and shifts from ideation, preregistration to execution of the plan. Two features define adaptive preregistration. It contains 'plans to deploy flexible strategies' \<Srivastava\> wherein the author can supply a heuristic consisting of multiple different analysis or modelling strategies whose execution depends on the outcome of previous decision-points or analyses. Secondly, the preregistration may be iterative and consist of interim preregistrations that mark phases of modelling and analysis as different parts of the data are observed \<Srivastava\>. Thus, as the modeller proceeds through the model development process, the researcher can shift from ideation, preregistration, execution of the analysis and back to ideation and preregistration again, depending on the observed outcomes as each interim preregistration is executed \<Link to Figure, GitHub as a platform for adaptive preregistration, <x-scrivener-item:///Users/egould/Dropbox/Apps/Scrivener/PhD_thesis.scriv?id=32FCFBDA-7CEE-4FD9-B8DD-F3073309C0D0>\>.

Given the iterative nature of adaptive preregistration then, the current convention of the static text-based document format of preregistration is inadequate. If there are interim versions of a preregistration, each version must be time-stamped and *version-controlled*, such that it is clear not just that two or more versions differ from each other, but it is apparent in exactly *how* they differ. Moreover, the *checking problem*, is exacerbated --- for flexible modelling and analysis strategies, we need to be able to link the results of preliminary analyses back to the specified strategy: the link between a particular triggered decision, and the analysis outcome that triggered it must be explicit. In addition, if the results of some preliminary analyses cause revisions to the model itself or a different modelling procedure or approach to be selected, then the trigger for this decision must be explicitly linked to the new version of the preregistration.

@@

I think the preregistration can easily be usd with a reporting check

It's, such as t proposed by Feng et al. The decisions and details are recorded within the preregistration. Along with any deviation from the initial plan.  
--------------------------
# References

>>>>

---
title: "Preregistration for modelling in ecolgy and conservation"
description: |
  Working Draft.
author:
  - name: Elliot Gould 
    affiliation: School of BioSciences, University of Melbourne
  - name: Chris Jones
    affiliation: Department of Environment Land and Water
  - name: Libby Rumpff
    affiliation: School of BioSciences, University of Melbourne
  - name: Hannah Fraser
    affiliation: School of BioSciences, University of Melbourne
  - name: Fiona Fidler
    affiliation: School of BioSciences, University of Melbourne
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    self_contained: false
---
**Intended journal:** Methods in Ecology and Evolution.

@@

**Title:** "But I can't preregister my research": Improving the reproducibility and transparency of ecology and conservation with preregistration for model-based research

**Abstract:** *Ecology is at risk of a 'reproducibility crisis', in particular, there are serious transparency problems in the reporting of ecological models, which is concerning because a lack of complete and transparent reporting practices provide the conditions in which questionable research practices are likely to arise, leading to irreproducible findings. Irreproducibility of model-based research threatens the progress of both scientific discovery in 'pure' research, and risks wrong or sub-optimal conservation and management decisions being implemented when models are used in decision-support. Preregistration is an open-science practice adopted in scientific disciplines that have confronted the reproducibility crisis, which aims to improve research transparency and mitigate questionable research practices. There has been little uptake of preregistration in ecology and conservation, and within disciplines wherein preregistration has been adopted and become common practice, many modellers and researchers engaged in non-hypothesis testing research have vehemently eschewed preregistration on the grounds that existing preregistration templates are irrelevant to their research because they are focussed on null-hypothesis significance testing. We took a user-centred design approach to the task of translating preregistration into model-based research contexts for ecology and conservation. Structured discussion and collaborative activities underpinned the development of a preregistration template that captures critical analysis decisions in the process of the model development cycle. The iterative nature of model development was identified as a key barrier to implementing preregistration for modelling research because it conflicts with the linear procedure of preregistration in which a researcher shifts from ideation to analysis planning, to analysis and then reporting. We propose a methodology for implementing 'adaptive preregistration' leveraging git and GitHub, which accommodates common practices in modelling where some analysis decisions are difficult or cannot be made in advance, where downstream decisions might be dependent on the outcome of earlier decisions, and where preliminary analyses might need to be conducted before proceeding with further modelling. We tested and evaluated the preregistration template and methodology using a case-study of environmental flows modelling for decision-support in Victoria, Australia. This research provides an extensible template and methodology for implementing adaptive preregistration of ecological models, to be used in either pure or applied settings. We hope that this research will facilitate wider uptake of preregistration in ecology and that our proposed methodology for implementing adaptive preregistration might be applied in model-based research within science more broadly. Modellers need no longer cry "but, I can't preregister my research."*

**Keywords** *(no more than 8 for MEE).***:** *preregistration, transparency, reproducibility, modelling, ecology, conservation, open-science, research software engineering.*

@@

Ecology as a discipline is at risk of a "reproducibility crisis" (Fidler et al. 2017), with rates of Questionable Research Practices (QRPs) in ecology similar to those in other fields (Fraser et al. 2018), such as psychology and medicine, where the crisis has received considerable attention, and provoked heated discussion among researchers within the broader scientific community. In particular, there are serious problems with the way ecological models and the modelling process are reported in the published literature. However, concerns about poor model reporting practices are not new, and have been raised repeatedly --- nearly a decade ago researchers concluded that most model reporting was neither transparent or comprehensive [@Schmolke:2010fd; @deVos:2011], and that a culture of model reporting was yet to be established [@Grimm:2014es]. Moreover, most models are assessed predominantly on the basis of scientific originality during peer-review  [@Schmolke:2010fd] to the effect that the burden of evaluating models for credibility is placed onto the reader or model-user, rather than the peer-review process, especially in the case of extremely complex models [@Alexandrov:2011iv].

Despite attempts to promote thorough and transparent model documentation using 'Good Modelling Practice' (e.g. [@Jakeman:2006ii;@Alexandrov:2011iv]), more recent quantitative assessments of transparency and reproducibility demonstrate that transparent model reporting is an ongoing problem. Just as Jakeman argued that many models in the published literature lack a clear statement of model objectives and purpose, lack descriptions of model assumptions and implications, and also fail to report model results, including validation and evaluation in 2006, [-@Jakeman:2006ii], [@Feng2019]'s quantitative assessment of Ecological Niche Models found that reporting of model evaluation and model results was less than adequate.

@@

@@

Figure X and Table X, below, describes the 'model scientific workflow' underpinning the template, which was identified through the structured collaborative workshop, literature review and qualitative coding. By articulating this workflow, we explicitly state the workflow and sets of activities that the template presumes. We identified and coded decision-points into 6 'phases' in the process of developing, evaluating and analysing a model or models in ecological application: 'problem formulation', 'define conceptual model', 'formalise and specify model', 'model validation and evaluation', 'model analysis', 'model application'. These phases are comprised of a series of smaller 'steps' and 'substeps', which are defined and described in detail within Table X. The broad phases and steps are used to structure the template by grouping similar modelling activities together, and the descriptions and definitions of the phases and steps are used as 'help-text' in the template interface.

@@

One of the key concerns raised at the workshop by those who work in applied contexts, is that preregistration could be a useful tool, but needed to consist of items that were useful to applied contexts. Another thing is that they felt that it could a) slot into existing paper-work and b) be used to make internal data analysis plan requirements more useful.

@@

During our structured discussions, we identified the iterative nature of model development as a significant barrier to adopting preregistration.  Translating the preregistration process to a modelling research context is thus challenging because the iterative nature of model development conflicts with the inner logic of existing preregistration templates that presume a linear research workflow. The process of preregistration as it is currently implemented is tailored towards the hypothetico-deductive model of the scientific method. As such, the process of preregistration is completely distinct from, and precedes the implementation of the analysis plan laid out in the preregistration. In fact, preregistration has been defined as "the action of confirming an unalterable version of one's research plan prior to collecting data" (Nosek et al. 2018) or prior to analysing the data [@Mertens2019]. The researcher shifts from ideation, to study/analysis design and preregistration, to collecting and analysing the data, writing the report or manuscript, to documenting the research and publishing ([*cos.io/rr*](http://cos.io/rr)) [See also Figure 1 in @Pu2019]. The iterative nature of model development, however, inherently precludes this model of preregistration from being applied to model-based research in ecology and conservation.

Firstly, as Srivastava notes [-@Srivastava2018], there are some decisions that are either difficult to anticipate, or simply cannot be made in advance. This is particularly true for decisions occurring at the later phases of the modelling construction and development process (Fig X), where downstream decisions depend on the outcomes of earlier decisions and outcomes of modelling analyses. Some items on the template might not be able to be answered on the first go by the researcher until the model is fully or at least close to fully specified, for example, specifying precisely what and how sensitivity analyses or uncertainty analyses will be conducted.

Secondly, very often in modelling, preliminary or investigatory analyses might need to be conducted before being able to specify future decision-steps in the analysis plan. For example, modellers might need to check the distribution of particular variables in order to determine how to specify the model, perform assumption checks, or check for collinearity or spatial autocorrelation. In addition, sometimes the results of these analyses and checks mean that the modeller is forced to return to earlier decision-points and change the analysis. For instance, the model might fail to converge, or is saturated and runs out of variation to apportion, such that the model must be respecified. Other times, the modelling process itself may generate knowledge and learning about the system, and the model structure changes throughout the process of model development.

In all of these examples, during the course of model-based preregistration a researcher would switch back and forth between preregistration and analysis, implementing parts of the analysis plan that were specified in the previous preregistration step, and letting the outcomes in that analysis step inform the next specification of the analysis. This contravenes a critical feature of preregistration as implemented in hypothetico-deductive frameworks, where the analysis plan must be fully specified prior to observing and analysing the data. This is not a flaw of existing preregistration, but rather reflects the inner-logic of preregistration for hypothesis-testing contexts. The procedure for implementing preregistration in modelling must therefore account for this iterative and non-linear feature of model development, should we wish to preregister model-based research.

@@

In order to address the disjuncture between the iterative nature of model development, and the single-use, deterministic preregistration template, we take an 'expanded view of preregistration', that fits with the concept of 'adaptive preregistration' proposed by \<Srivastava\>. This view of preregistration breaks with the current model of preregistration, wherein the author writes a single deterministic preregistration containing a rule for every decision, and shifts from ideation, preregistration to execution of the plan. Two features define adaptive preregistration. It contains 'plans to deploy flexible strategies' \<Srivastava\> wherein the author can supply a heuristic consisting of multiple different analysis or modelling strategies whose execution depends on the outcome of previous decision-points or analyses. Secondly, the preregistration may be iterative and consist of interim preregistrations that mark phases of modelling and analysis as different parts of the data are observed \<Srivastava\>. Thus, as the modeller proceeds through the model development process, the researcher can shift from ideation, preregistration, execution of the analysis and back to ideation and preregistration again, depending on the observed outcomes as each interim preregistration is executed \<Link to Figure, GitHub as a platform for adaptive preregistration, <x-scrivener-item:///Users/egould/Dropbox/Apps/Scrivener/PhD_thesis.scriv?id=32FCFBDA-7CEE-4FD9-B8DD-F3073309C0D0>\>.

The iterative nature of model construction and development complicates what is known as the preregistration *checking problem* even further. Although there are several aspects to the *checking problem*, here we focus on the part of the checking problem that includes *how do we facilitate checking?* Authors, reviewers, editors and readers may all engage in the process of checking a manuscript or published paper against its preregistration to verify that the study and analyses were conducted as specified in the preregistration. Existing pre-registration formats hinder checking, because the prevailing format of a single static text-based document is designed for authors to very quickly input the required information, rather than to facilitate a comparison of the reported analysis and results against the planned analysis within the preregistration. In fact, *Pu et al* go so far as to say that this format "verge[s] on being write-only media" and are therefore completely ill-suited to the checking task. The task of model checking is further complicated by the non-linear and iterative nature of model development, and by our proposed adaptive preregistration methodology because in most modelling studies there will likely be interim preregistrations, such that there are *multiple* versions of the preregistration to check against the completed analysis.

In redesigning the procedure and interface for completing a preregistration using an adaptive preregistration process, we followed the advice by *Pu et al. [-@Pu2019]* who recommended two design elements that may facilitate checking. Firstly, that the design should "make relevant preregistration easier to *query"* and secondly it should "encourage wider *coverage* of preregistration content during review." The process of *querying* a preregistration consists of "paper-to-preregistration" and "preregistration-to-paper" lookup whereby the reader searches for relevant information from the paper in the preregistration and *vice-versa*. This requires that the report or paper must be specific in referring to parts of the preregistration, *and*, that the author has tagged the relevant part of the preregistration in their report or manuscript.

Given the iterative nature of adaptive preregistration then, the current convention of the single-use, deterministic and static text-based document format of preregistration is inadequate. Interim versions of a preregistration must be time-stamped and ideally *version-controlled*, such that it is clear not just that two or more versions differ from each other, and in what order they were created, but so that it is explicitly clear in how they differ. To facilitate preregistration checking of flexible modelling and analysis strategies, we must link the results of both final and preliminary analyses back to the specified strategy: the link between a particular triggered decision, and the analysis outcome that triggered the decision must be explicit. Additionally, if the results of some preliminary analyses cause revisions to the model itself or a different modelling procedure or approach to be selected than was originally or previously planned, then the trigger for this decision must be explicitly linked to the next version of the preregistration.

Consequently we propose a methodology for implementing adaptive preregistration that leverages features of GitHub to also facilitate the task of linking the actual study and analysis conducted to the pre-specified plan in the preregistration \<link to part of GH methodology that explains how checking is accounted for in the procedural design\>. In addition, we designed a shiny interface for completing the preregistration to include a 'checking mode' that easily allows the author to state how and why the the analysis might have diverged from the planned \<link to part of shiny app interface that discusses checking mode\>.

@@

@@

I thinkthe preregistration can easily be usd with a reporting check

It's, such as t proposed by Feng et al. The decisions and details are recorded within the preregistration. Along with any deviation from the initial plan.

**Continuous Improvement and Feedback**

One of the key features of user-centred design is the notion of 'continuous feedback'.

We encourage modellers to preregister their research using our template and methodology here, and to provide feedback and suggestions for improvement. Our template + shiny app is hosted on \<insert GitHub link\>, suggested changes may be submitted in the form of a pull-request to the repository \<link to how to make a pull request\>, or by posting a new issue \<link to creating a new issue on the repo\>. The repository and issues are publicly available, so any discussion will be transparent and open to the ecological modelling community.  
--------------------------
# References
